
---
title: "CIMMS PSET 2 — Answer Key"
author: "Answer Key"
output: html_document
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)
library(janitor)
```

```{r load-data}
df_raw <- readr::read_csv("healthcare_messy_data.csv", show_col_types = FALSE)
```

# Part A — First Look

## A1. Basic Structure

```{r}
head(df_raw, 10)
glimpse(df_raw)
dim(df_raw)
```

## A2. Quick Counts

```{r}
df_raw %>%
  count(Gender)

df_raw %>%
  count(Condition, .drop = FALSE)
```

## A3. Cholesterol Average

```{r}
df_raw %>%
  filter(Cholesterol != "NaN") %>%
  mutate(Cholesterol = as.numeric(Cholesterol)) %>%
  summarise(mean_chol = mean(Cholesterol, na.rm = TRUE))

df_raw %>%
  filter(Cholesterol != "NaN") %>%
  mutate(Cholesterol = as.numeric(Cholesterol)) %>%
  group_by(Gender) %>%
  summarise(mean_chol = mean(Cholesterol, na.rm = TRUE))
```

# Part B — Cleaning

## B1. Standardize Column Names

```{r}
df_clean <- df_raw %>%
  clean_names()
```

## B2. Convert Dates

```{r}
df_clean <- df_clean %>%
  mutate(
    visit_date = parse_date_time(visit_date, orders = c("mdy", "ymd", "dmy"))
  )
```

## B3. Clean Gender

```{r}
df_clean <- df_clean %>%
  mutate(
    gender = str_to_lower(gender),
    gender = case_when(
      gender %in% c("m", "male") ~ "Male",
      gender %in% c("f", "female") ~ "Female",
      TRUE ~ NA_character_
    )
  )
```

## B4. Convert Blood Pressure

```{r}
df_clean <- df_clean %>%
  separate(blood_pressure, into = c("systolic", "diastolic"), sep = "/", convert = TRUE)
```

# Part C — Transformation

## C1. BMI

```{r}
df_clean <- df_clean %>%
  mutate(bmi = weight_kg / (height_cm/100)^2)
```

## C2. Age Groups

```{r}
df_clean <- df_clean %>%
  mutate(age_group = case_when(
    age < 30 ~ "<30",
    age >= 30 & age < 50 ~ "30-49",
    age >= 50 ~ "50+"
  ))
```

## C3. Mean Blood Pressure

```{r}
df_clean %>%
  group_by(condition) %>%
  summarise(
    mean_sys = mean(systolic, na.rm = TRUE),
    mean_dia = mean(diastolic, na.rm = TRUE)
  )
```

# Part D — Summaries

## D1. Visit Counts

```{r}
df_clean %>%
  count(patient_id)
```

## D2. Long Format

```{r}
df_long <- df_clean %>%
  pivot_longer(
    cols = c(cholesterol, bmi, systolic, diastolic),
    names_to = "measure",
    values_to = "value"
  )
```

## D3. Condition vs Gender

```{r}
df_clean %>%
  tabyl(condition, gender) %>%
  adorn_totals("both")
```

# Part E — Visualization

## E1. Cholesterol Distribution

```{r}
ggplot(df_clean, aes(x = cholesterol)) +
  geom_histogram(bins = 30)
```

## E2. BMI by Condition

```{r}
ggplot(df_clean, aes(x = condition, y = bmi)) +
  geom_boxplot()
```

## E3. Visits Over Time

```{r}
df_clean %>%
  mutate(month = floor_date(visit_date, "month")) %>%
  count(month) %>%
  ggplot(aes(x = month, y = n)) +
  geom_line()
```
